% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={On One-Dimensional Quantum Cellular Automata },
  pdfauthor={John Watrous* Computer Sciences Department University of Wisconsin Madison, Wisconsin 53706},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{On One-Dimensional Quantum Cellular Automata}
\author{John Watrous*\\
Computer Sciences Department\\
University of Wisconsin\\
Madison, Wisconsin 53706}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Since Richard Feynman introduced the notion of quantum computation in
1982, various models of "quantum computers" have been proposed. These
models include quantum Turing machines and quantum circuits. In this
paper we define another quantum computational model, one-dimensional
quantum cellular automata, and demonstrate that any quantum Turing
machine can be efficiently simulated by a onedimensional quantum
cellular automaton with constant slowdown. This can be accomplished by
consideration of a restricted class of one-dimensional quantum cellular
automata called one-dimensional partitioned quantum cellular automata.
We also show that any one-dimensional partitioned quantum cellular
automaton can be simulated by a quantum Turing machine with linear
slowdown, but the problem of efficiently simulating an arbitrary
one-dimensional quantum cellular automaton with a quantum Turing machine
is left open. From this discussion, some interesting facts concerning
these models are easily deduced.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The idea that certain principles of quantum mechanics might be powerful
computational tools has led to the study of various theoretical models
of quantum computers. The archetypal quantum computer, as introduced by
Richard Feynman in {[}1{]}, is a computer which can simulate quantum
physical processes, and which operates in accordance with quantum
physical laws. Feynman noted that the problem of simulating quantum
physics with a computer based on classical physics appears to be
intractable, thereby suggesting that quantum computers may be inherently
more powerful than classical computers. David Deutsch later formalized
the notion of quantum computation by defining what has become known as
the

\begin{itemize}
\item
  Supported in part by NSF Grant CCR-9208639. quantum Turing machine
  (QTM) in {[}2{]}, and in {[}3{]} Ethan Bernstein and Umesh Vazirani,
  expanding on Deutsch's work, showed that there exists a universal QTM
  which can simulate any QTM to any required accuracy with at most
  polynomial slowdown. Another quantum computational model, the quantum
  circuit, was introduced by Andrew Yao in {[}4{]} and shown to be
  equivalent to the quantum Turing machine.
\end{itemize}

It is not known whether or not these models are more powerful than their
classical analogues. However, there is evidence to suggest that this is
the case; most notably Peter Shor has shown in {[}5{]} that the integer
factoring and discrete log problems can be solved in polynomial time
using a QTM. (These problems are believed not to be solvable in
polynomial time using a probabilistic Turing machine.)

It is natural to extend the idea of quantum computation to other
computational models; in this paper we discuss one-dimensional quantum
cellular automata. Given any well-formed QTM, we give a construction of
a one-dimensional quantum cellular automaton (1dQCA) which will
efficiently simulate this QTM. This is accomplished by defining a
restricted class of 1dQCA called one-dimensional partitioned quantum
cellular automata (1d-PQCA). A 1d-PQCA is a 1d-QCA in which each cell is
partitioned into three subcells (left, middle, and right), and where the
next states of any given cell depend only on the contents of the right
subcell of its left neighbor, the middle subcell of itself, and the left
subcell of its right neighbor. This is the quantum analogue of the
partitioned cellular automaton discussed by Kenichi Morita and Masateru
Harao in {[}6{]}. The advantage of the 1d-PQCA class is that it is a
simple matter to determine whether or not a given 1d-PQCA is
well-formed, while this is not a trivial matter for an arbitrary 1d-QCA.

It is not clear that an arbitrary \(1 \mathrm{~d}-\mathrm{QCA}\) can be
efficiently simulated by a QTM. However, it is shown that any given
1d-PQCA can be simulated by a QTM with linear slowdown. It is
interesting to note that given any 1d-PQCA, the QTM which results from
this con- struction will simulate the given 1d-PQCA with deterministic
head position, i.e. if this QTM is observed at any time during its
computation, the probability that the tape head will be observed in any
given location will be either 0 or 1 . This allows for the construction
of a QTM which will simulate, with deterministic head position, any
given QTM with linear slowdown. Thus, for example, the position of the
tape head of any such QTM could be observed at every time step without
affecting its computation. (This is generally not the case for an
arbitrary QTM.)

The remainder of this paper will be organized as follows. In section 2,
the one-dimensional quantum cellular automata model is defined. In
section 3, the class of one-dimensional partitioned quantum cellular
automata is defined, and necessary and sufficient conditions for the
well-formedness of a 1d-PQCA are discussed. In section 4 , the quantum
Turing machine model is reviewed and in section 5 , the equivalence of
the quantum Turing machine model and the partitioned quantum cellular
automata model is demonstrated. Finally, in section 6 , some facts
resulting from this discussion are mentioned.

\hypertarget{one-dimensional-quantum-cellular-automata}{%
\section{One-Dimensional Quantum Cellular
Automata}\label{one-dimensional-quantum-cellular-automata}}

A one-dimensional quantum cellular automaton \(M\) is a quadruple
\((Q, \delta, k, A)\) where \(Q\) is a finite set of states (including a
distinguished quiescent state denoted by \(\epsilon\) ), \(\delta\) is a
local transition function (described below), \(k\) is an integer
denoting the acceptance cell, and \(A \subseteq Q\) is a set of
accepting states. \(M\) is assumed to have a two-way infinite sequence
of cells indexed by the integers (hereafter denoted by \(\mathbb{Z}\).)
The neighborhood of each cell is defined to be that cell itself along
with its closest neighbor on each side.

A configuration of a 1d-QCA \(M\) is a map

\[a: \mathbb{Z} \rightarrow Q\]

where, for each integer \(n, a(n)\) denotes the state of the cell
indexed by \(n\). For any configuration \(a\), it is assumed that there
are only finitely many values of \(n\) for which \(a(n)\) is a
non-quiescent state. Denote by \(\mathcal{C}=\mathcal{C}(M)\) the set of
all configurations of \(M\), so that \(\mathcal{C}\) is countable for
any given 1d-QCA \(M\). For any \(a \in \mathcal{C}\), define \(|a|\)
(the length of \(a\) ) to be the maximum number of consecutive cells
such that the first and last cells are non-quiescent. Any configuration
of \(M\) in which the cell indexed by \(k\) contains an element of \(A\)
is said to be an accepting configuration and all other configurations
are non-accepting configurations.

The local transition function \(\delta\) is a map

\[\delta: Q^{4} \longrightarrow \mathbb{C}\]

with

\[\delta(\epsilon, \epsilon, \epsilon, q)= \begin{cases}1 & \text { if } q=\epsilon \\ 0 & \text { if } q \neq \epsilon\end{cases}\]

which describes the evolution of \(M\). (Here \(\mathbb{C}\) denotes the
field of complex numbers.) Suppose that at some given time \(t\), three
consecutive cells of \(M\) are in states \(q_{1}, q_{2}\) and \(q_{3}\)
respectively. Then for every state \(q \in Q\), the cell which contained
\(q_{2}\) at time \(t\) will, at time \(t+1\), update to the state \(q\)
with amplitude \(\delta\left(q_{1}, q_{2}, q_{3}, q\right)\). Every cell
updates simultaneously in this manner, so that globally, if \(M\) is in
some configuration \(a \in \mathcal{C}\) at time \(t\), then the
amplitude with which \(M\) transforms to any configuration
\(b \in \mathcal{C}\) at time \(t+1\) is defined to be the product of
the amplitudes with which each cell of \(a\) transforms to the
corresponding cell of \(b\), i.e.

\[\prod_{n \in \mathbb{Z}} \delta(a(n-1), a(n), a(n+1), b(n)) .\]

(This product is guaranteed to exist by (1).)

Thus, any configuration of \(M\) will transform into multiple "next
configurations", where the transformation to each individual
configuration has an associated amplitude. The evolution behaves as if
all of these transformations occur simultaneously, so that after some
number of steps, \(M\) will have evolved along many computation paths
simultaneously. Each path has an associated amplitude which is defined
to be the product of the amplitudes of the transformations along that
path.

Multiple paths with differing amplitudes may lead from one given
configuration to another. This effect is known as interference. If \(M\)
is assumed to be in some configuration \(a\) and is allowed to evolve
for, say \(l\) steps, then \(M\) will be in a linear superposition of
configurations. For each configuration \(b\) in such a superposition, we
associate with it an amplitude which is the sum of the amplitudes of all
paths of length \(l\) from \(a\) to \(b\). It may therefore be the case,
for example, that a configuration will have amplitude zero in some
superposition, despite the fact that multiple paths of nonzero amplitude
lead to it.

If a 1d-QCA \(M\) is observed while in some superposition of
configurations, the observer will not see this superposition. Rather,
the act of observation forces the machine to choose one of the
configurations in the given superposition randomly, so that exactly one
configuration will be observed. The probability that any given
configuration is observed is the absolute square of the amplitude
associated with that configuration, i.e. the absolute square of the sum
of the amplitudes of all paths which lead to that configuration. The act
of observation has the effect of altering the machine, so that
immediately after the observation occurs the machine will be in a single
configuration, and no longer a superposition of configurations. (More
general types of observations are possible, but will be ignored for the
purposes of this paper.)

For any 1d-QCA \(M\) and input configuration \(a\), we are interested in
whether or not an accepting configuration will be observed after some
number of steps. The probability that \(M\) accepts a after \(l\) steps
is simply the probability of observing any accepting configuration after
\(l\) steps, assuming that \(M\) is initially placed in configuration
\(a\) and is not observed before \(l\) steps have passed.

Since each configuration in a given superposition is observed with a
certain probability, it is necessary that the sum of the probabilities
be exactly one. Thus, allowed local transition functions \(\delta\) must
be restricted to those that guarantee this condition for any
superposition resulting from any given input. A machine with such a
local transition function is said to be wellformed. This will be
formalized presently.

Given any 1d-QCA \(M\), let \(\ell_{2}(\mathcal{C})\) denote the space
of all complex valued functions with domain
\(\mathcal{C}=\mathcal{C}(M)\) and bounded \(\ell_{2}\)-norm, i.e.

\[\ell_{2}(\mathcal{C})=\left\{x: \mathcal{C} \rightarrow \mathbb{C} \mid\left(\sum_{a \in \mathcal{C}} x(a) \overline{x(a)}\right)^{1 / 2}<\infty\right\} .\]

Then \(\ell_{2}(\mathcal{C})\) is a Hilbert space with respect to the
inner product
\(\langle\cdot, \cdot\rangle: \ell_{2}(\mathcal{C}) \times \ell_{2}(\mathcal{C}) \rightarrow \mathbb{C}\)
defined by

\[\left\langle x_{1}, x_{2}\right\rangle=\sum_{a \in \mathcal{C}} x_{1}(a) \overline{x_{2}(a)} .\]

Any superposition of \(M\) can now be identified with an element
\(x \in \ell_{2}(\mathcal{C})\), where \(x(a) \in \mathbb{C}\) denotes
the amplitude associated with configuration \(a\) in this superposition.
Thus, the probability of observing \(a\) from this superposition is
\(|x(a)|^{2}\) for any \(a \in \mathcal{C}\). We must therefore have

\[\sum_{a \in \mathcal{C}}|x(a)|^{2}=1\]

for any superposition \(x\). The sum in (2) is exactly \(\|x\|^{2}\) by
definition, so we define

\[\mathcal{S}=\left\{x \in \ell_{2}(\mathcal{C}) \mid\|x\|=1\right\}\]

to be the set of all possible superpositions of \(M\). Each
superposition of a given 1d-QCA \(M\) will, in one time-step, evolve
into a new superposition according to the local transition function of
\(M\). We can associate with any \(M\) a function \(E\) which will map
any given superposition to this next superposition; if \(x\) is some
superposition of \(M\) at time \(t\), then \(M\) will be in the
superposition \(E x\) at time \(t+1\). In general, after \(l\) steps,
\(M\) will be in the superposition \(E^{l} x\). \(E\) is the
time-evolution operator of \(M\), and can be explicitly defined as
follows. For all \(a, b \in \mathcal{C}\) let \(\alpha(a, b)\) denote
the amplitude with which configuration \(a\) transforms into
configuration \(b\), i.e.

\[\alpha(a, b)=\prod_{n \in \mathbb{Z}} \delta(a(n-1), a(n), a(n+1), b(n)),\]

and for each \(x \in \ell_{2}(\mathcal{C})\) and \(b \in \mathcal{C}\)
define

\[\operatorname{Ex}(b)=\sum_{a \in \mathcal{C}} \alpha(a, b) x(a)\]

Thus, we say that \(M=(Q, \delta, k, A)\) is a well formed
one-dimensional quantum cellular automaton (write
\(M \in 1 \mathrm{~d}-\mathrm{QCA}\) ) if and only if the corresponding
timeevolution operator \(E\) preserves \(\ell_{2}\)-norm, i.e.

\[x \in \mathcal{S} \Longleftrightarrow E x \in \mathcal{S} .\]

\hypertarget{partitioned-quantum-cellular-au--tomata}{%
\section{Partitioned Quantum Cellular Au-
tomata}\label{partitioned-quantum-cellular-au--tomata}}

In general, given an arbitrary \(M=(Q, \delta, k, A)\), it is not a
trivial matter to determine whether or not \(M\) is well formed. For
this reason we now define a restricted class of
\(1 \mathrm{~d}-\mathrm{QCA}\) called partitioned quantum cellular
automata for which this can easily be determined. This is a
generalization of (deterministic) partitioned cellular automata
discussed by Morita and Harao in {[}6{]}.

A one-dimensional partitioned quantum cellular automaton is a 1d-QCA in
which each cell is partitioned into three subcells: a left subcell, a
middle subcell and a right subcell. (The set of states \(Q\) is
decomposed accordingly.) The next state(s) of any cell may now only
depend upon the states of the left subcell of the right neighbor, the
middle subcell of the cell itself, and the right subcell of the left
neighbor.

More formally, let \(M=(Q, \delta, k, A)\) as in the \(1 \mathrm{~d}\)
QCA case, where \(Q\) and \(\delta\) are restricted as follows. Let

\[Q=Q_{l} \times Q_{m} \times Q_{r}\]

for finite sets \(Q_{l}, Q_{m}\) and \(Q_{r}\), (and again let
\(\epsilon\) denote the distinguished quiescent element of \(Q\).) Let
\(\Lambda\) be an \(|Q| \times|Q|\) matrix over \(\mathbb{C}\) having
the form

\[\Lambda=\left(\begin{array}{cccc}
\lambda\left(q_{1}, q_{1}\right) & \lambda\left(q_{1}, q_{2}\right) & \ldots & \lambda\left(q_{1}, q_{|Q|}\right) \\
\lambda\left(q_{2}, q_{1}\right) & \lambda\left(q_{2}, q_{2}\right) & \ldots & \lambda\left(q_{2}, q_{|Q|}\right) \\
\vdots & \vdots & \ddots & \vdots \\
\lambda\left(q_{|Q|}, q_{1}\right) & \lambda\left(q_{|Q|}, q_{2}\right) & \ldots & \lambda\left(q_{|Q|}, q_{|Q|}\right)
\end{array}\right)\]

where \(\lambda: Q \times Q \longrightarrow \mathbb{C}\) must satisfy

\[\lambda(\epsilon, q)=\lambda(q, \epsilon)= \begin{cases}1 & \text { if } q=\epsilon \\ 0 & \text { if } q \neq \epsilon .\end{cases}\]

For any state \(q=\left(q_{l}, q_{m}, q_{r}\right)\) define

\[l(q)=q_{l}, \quad m(q)=q_{m}, \quad r(q)=q_{r},\]

and define \(\delta\) as

\[\delta\left(q_{1}, q_{2}, q_{3}, q\right)=\lambda\left(\left(l\left(q_{3}\right), m\left(q_{2}\right), r\left(q_{1}\right)\right), q\right)\]

for all \(q_{1}, q_{2}, q_{3}, q \in Q\). Since a given matrix
\(\Lambda\) completely determines \(\delta\) in this manner, we may
write \(M=(Q, \Lambda, k, A)\) rather than \(M=(Q, \delta, k, A)\) for a
1d-PQCA \(M\) whenever convenient.

The behavior of a partitioned quantum cellular automaton \(M\) is, in
general, simpler than that of an arbitrary 1d-QCA. Suppose that \(M\) is
in configuration \(a\). Then after one time step, for any given
\(n \in \mathbb{Z}\), the state of the cell indexed by \(n\) will be
transformed to each state \(q\) with amplitude

\[\begin{aligned}
& \delta(a(n-1), a(n), a(n+1), q) \\
& \quad=\lambda((l(a(n+1)), m(a(n)), r(a(n-1))), q) .
\end{aligned}\]

In order to simplify this, we will define a permutation
\(\sigma: \mathcal{C} \longrightarrow \mathcal{C}\) as

\[\sigma a(n)=(l(a(n+1)), m(a(n)), r(a(n-1)))\]

for all \(n \in \mathbb{Z}\). The action of the permutation \(\sigma\)
on any configuration \(a\) can be illustrated as follows:

{image}

Now, (4) is equivalent to

\[\delta(a(n-1), a(n), a(n+1), q)=\lambda(\sigma a(n), q),\]

so that applying \(\delta\) to each neighborhood of some configuration
\(a\) is equivalent to first applying \(\sigma\) to \(a\) and then
applying \(\lambda\) to each individual cell of \(\sigma a\), i.e.
\(\sigma a(n)\) is transformed into each state \(q\) with amplitude
\(\lambda(\sigma a(n), q)\) for every \(n \in \mathbb{Z}\).

If, for such an \(M\), we have \(M \in 1 \mathrm{~d}-\mathrm{QCA}\) we
will say that \(M\) is a well formed one-dimensional partitioned quantum
cellular automaton (write \(M \in 1\) d-PQCA.) The following theorem
allows us to characterize onedimensional partitioned cellular automata
in terms of the matrix \(\Lambda\).

Theorem 3.1 Let \(M=(Q, \Lambda, k, A)\) with \(\Lambda\) as above. Then
\(M \in 1 d-P Q C A\) if and only if \(\Lambda\) is unitary.

The remainder of this section will be devoted to proving this theorem.

For each \(a, b \in \mathcal{C}\) define

\[u_{b}(a)=\overline{\alpha(a, b)}\]

and

\[v_{b}(a)=\alpha(b, a) .\]

Lemma 3.1 Let \(M=(Q, \Lambda, k, A)\), where \(\Lambda\) satisfies (3).
Then we have

\[\#\left\{a \in \mathcal{C} \mid u_{b}(a) \neq 0\right\}<\infty\]

and

\[\#\left\{a \in \mathcal{C} \mid v_{b}(a) \neq 0\right\}<\infty\]

for any \(b \in \mathcal{C}\).

Proof. Given \(b \in \mathcal{C}\), assume that \(u_{b}(a) \neq 0\). By
definition

\[u_{b}(a)=\prod_{n \in \mathbb{Z}} \overline{\lambda(\sigma a(n), b(n))} .\]

Since \(b \in \mathcal{C}\) and \(\lambda(q, \epsilon)=0\) for
\(q \neq \epsilon, a\) must be such that \(\sigma a(n)=b(n)=\epsilon\)
for all but finitely many \(n\). There are finitely many such
\(a \in \mathcal{C}\).

(6) follows similarly.

By Lemma 3.1, the sums

\[\sum_{a \in \mathcal{C}} u_{b}(a) \overline{u_{b}(a)}\]

and

\[\sum_{a \in \mathcal{C}} v_{b}(a) \overline{v_{b}(a)}\]

contain only finitely many nonzero terms and therefore converge, so we
have that \(u_{b}\) and \(v_{b}\) are elements of
\(\ell_{2}(\mathcal{C})\) for each \(b \in \mathcal{C}\). Lemma 3.2
\(\Lambda\) unitary
\(\Longrightarrow\left\{u_{b}\right\}_{b \in \mathcal{C}}\) and
\(\left\{v_{b}\right\}_{b \in \mathcal{C}}\) are orthonormal sequences
in \(\ell_{2}(\mathcal{C})\).

Proof. Given \(a, b \in \mathcal{C}\) we will show

\[\left\langle u_{a}, u_{b}\right\rangle= \begin{cases}1 & \text { if } a=b \\ 0 & \text { if } a \neq b\end{cases}\]

For each \(m \in \mathbb{Z}, q \in Q\), define
\(S_{m}^{q}: \mathcal{C} \longrightarrow \mathcal{C}\) as

\[S_{m}^{q}(c)(n)= \begin{cases}c(n) & \text { if } n>m \\ q & \text { if } n=m \\ c(n+1) & \text { if } n<m\end{cases}\]

for all \(n \in \mathbb{Z}\), and also define
\(R_{m}^{q}: \mathcal{C} \longrightarrow \mathcal{C}\) as

\[R_{m}^{q}(c)(n)= \begin{cases}c(n) & \text { if } n \neq m \\ q & \text { if } n=m\end{cases}\]

for all \(n \in \mathbb{Z}\).

Since \(\sigma\) is one-to-one and onto, we have

\[\begin{aligned}
\left\langle u_{a}, u_{b}\right\rangle & =\sum_{c \in \mathcal{C}} u_{a}\left(\sigma^{-1} c\right) \overline{u_{b}\left(\sigma^{-1} c\right)} \\
& =\sum_{c \in \mathcal{C}} \prod_{n \in \mathbb{Z}} \overline{\lambda(c(n), a(n))} \lambda(c(n), b(n)) .
\end{aligned}\]

By Lemma 3.1 there are only finitely many values of \(c\) for which the
summand is nonzero, so we are free to change the order of summation in
(8) as follows

\[\left\langle u_{a}, u_{b}\right\rangle=\sum_{c \in \mathcal{C}} \sum_{q \in Q} P \overline{\lambda(q, a(m))} \lambda(q, b(m))\]

where

\[P=\prod_{n \neq m} \overline{\lambda\left(S_{m}^{q}(c)(n), a(n)\right)} \lambda\left(S_{m}^{q}(c)(n), b(n)\right)\]

and \(m\) is any integer. \(P\) is independent of \(q\), so we have

\[\left\langle u_{a}, u_{b}\right\rangle=\sum_{c \in \mathcal{C}} P \sum_{q \in Q} \overline{\lambda(q, a(m))} \lambda(q, b(m)) .\]

Suppose that \(a \neq b\). Then there must exist \(m \in \mathbb{Z}\)
such that \(a(m) \neq b(m)\). \(\Lambda\) is unitary, so that

\[\sum_{q \in Q} \overline{\lambda(q, a(m))} \lambda(q, b(m))=0\]

and therefore by (9) we have
\(\left\langle u_{a}, u_{b}\right\rangle=0\).

Now consider \(\left\langle u_{a}, u_{a}\right\rangle\). From (9)

\[\left\langle u_{a}, u_{a}\right\rangle=\sum_{c \in \mathcal{C}} P \sum_{q \in Q}|\lambda(q, a(m))|^{2}\]

for

\[P=\prod_{n \neq m}\left|\lambda\left(S_{m}^{q}(c)(n), a(n)\right)\right|^{2}\]

For any \(p \in Q\) we have \(\sum_{q \in Q}|\lambda(p, q)|^{2}=1\),
since \(\Lambda\) is unitary. Thus

\[\begin{aligned}
\left\langle u_{a}, u_{a}\right\rangle & =\sum_{c \in \mathcal{C}} \prod_{n \neq m}\left|\lambda\left(S_{m}^{q}(c)(n), a(n)\right)\right|^{2} \\
& =\sum_{c \in \mathcal{C}} \prod_{n \neq m}\left|\lambda\left(S_{m}^{\epsilon}(c)(n), a(n)\right)\right|^{2} \\
& =\sum_{c \in \mathcal{C}} \prod_{n \in \mathbb{Z}}\left|\lambda\left(c(n), R_{m}^{\epsilon}(a)(n)\right)\right|^{2} \\
& =\left\langle u_{a_{1}}, u_{a_{1}}\right\rangle
\end{aligned}\]

for \(a_{1}=R_{m}^{\epsilon}(a)\).

Now \(a \in \mathcal{C}\), and therefore there exists a finite set
\(\left\{m_{1}, \ldots, m_{k}\right\}\) for which

\[n \notin\left\{m_{1}, \ldots, m_{k}\right\} \Longrightarrow a(n)=\epsilon .\]

If we repeat the process resulting in (10) for each
\(m=m_{1}, m_{2}, \ldots, m_{k}\), we get
\(\left\langle u_{a}, u_{a}\right\rangle=\left\langle u_{a_{k}}, u_{a_{k}}\right\rangle\)
for
\(a_{k}=R_{m_{1}}^{\epsilon}\left(R_{m_{2}}^{\epsilon}\left(\cdots\left(R_{m_{k}}^{\epsilon}(a)\right) \cdots\right)\right)\).
Since \(a_{k}(n)=\epsilon\) for all \(n \in \mathbb{Z}\) we have

\[\begin{aligned}
\left\langle u_{a}, u_{a}\right\rangle & =\left\langle u_{a_{k}}, u_{a_{k}}\right\rangle \\
& =\sum_{c \in \mathcal{C}} \prod_{n \in \mathbb{Z}}|\lambda(c(n), \epsilon)|^{2}=1,
\end{aligned}\]

and so we have shown (7).

A slight modification of the above argument shows that
\(\left\{v_{b}\right\}_{b \in \mathcal{C}}\) is also an orthonormal
sequence.

Proof of Theorem 3.1. Assume that we are given \(M=(Q, \Lambda, k, A)\)
with \(\Lambda\) unitary. We will show that \(E\), the time-evolution
operator of \(M\), is unitary.

By definition, we have

\[E x(b)=\sum_{a \in \mathcal{C}} x(a) \overline{u_{b}(a)} .\]

so that

\[E x(b)=\left\langle x, u_{b}\right\rangle\]

for every \(x \in \ell_{2}(\mathcal{C})\) and \(b \in \mathcal{C}\).
Also define \(F\) as

\[F x(b)=\sum_{a \in \mathcal{C}} x(a) \overline{v_{b}(a)},\]

so that

\[F x(b)=\left\langle x, v_{b}\right\rangle\]

for every \(x \in \ell_{2}(\mathcal{C})\) and \(b \in \mathcal{C}\). By
Lemma 3.1, \(\left\{u_{b}\right\}_{b \in \mathcal{C}}\) is an
orthonormal sequence. So, by Bessel's inequality

\[\begin{aligned}
\|E x\|^{2} & =\sum_{b \in \mathcal{C}} \operatorname{Ex}(b) \overline{\operatorname{Ex}(b)} \\
& =\sum_{b \in \mathcal{C}}\left|\left\langle x, u_{b}\right\rangle\right|^{2} \\
& \leq\|x\|^{2},
\end{aligned}\]

for any \(x \in \ell_{2}(\mathcal{C})\). Thus \(E\) is a bounded linear
operator, and hence we have that

\[\langle E x, y\rangle=\sum_{b \in \mathcal{C}} \sum_{a \in \mathcal{C}} x(a) \overline{u_{b}(a) y(b)}\]

is a bounded bilinear form. This allows us (see {[}7{]}, for example) to
change the order of summation in (11) to get

\[\begin{aligned}
\langle E x, y\rangle & =\sum_{a \in \mathcal{C}} \sum_{b \in \mathcal{C}} x(a) \overline{u_{b}(a) y(b)} \\
& =\sum_{a \in \mathcal{C}} x(a) \sum_{b \in \mathcal{C}} \overline{y(b)} v_{a}(b) \\
& =\langle x, F y\rangle
\end{aligned}\]

for any \(x, y \in \ell_{2}(\mathcal{C})\), so that \(E^{*}=F\).

Now, for any \(x \in \ell_{2}(\mathcal{C})\) and \(b \in \mathcal{C}\)
we have, by applying Lemma 3.1 and Lemma 3.2,

\[\begin{aligned}
E^{*} \operatorname{Ex}(b) & =\sum_{a \in \mathcal{C}} \operatorname{Ex}(a) \overline{v_{b}(a)} \\
& =\sum_{a \in \mathcal{C}} \sum_{c \in \mathcal{C}} x(c) \overline{u_{a}(c) v_{b}(a)} \\
& =\sum_{c \in \mathcal{C}} \sum_{a \in \mathcal{C}} x(c) v_{c}(a) \overline{v_{b}(a)} \\
& =\sum_{c \in \mathcal{C}} x(c)\left\langle v_{c}, v_{b}\right\rangle \\
& =x(b)
\end{aligned}\]

so that \(E^{*} E=I\). Similarly

\[E E^{*} x(b)=\sum_{c \in \mathcal{C}} x(c)\left\langle u_{c}, u_{b}\right\rangle=x(b)\]

for any \(x \in \ell_{2}(\mathcal{C}), b \in \mathcal{C}\), so that
\(E E^{*}=I\).

Thus, \(E\) is unitary and hence preserves \(\ell_{2}\)-norm. We
therefore have that \(M \in 1 \mathrm{~d}-\mathrm{PQCA}\).

The converse is straightforward.

Corollary 3.1 For any \(M \in 1 d-P Q C A\), the associated
time-evolution operator \(E\) is unitary.

\hypertarget{quantum-turing-machines}{%
\section{Quantum Turing Machines}\label{quantum-turing-machines}}

In this section we will review the quantum Turing machine model as
defined in {[}3{]}. A quantum Turing machine \(M\) is a quintuple
\((K, \Sigma, \mu, k, A)\) where \(K\) is a finite set of states,
\(\Sigma\) is a finite tape alphabet (including a distinguished blank
symbol, denoted by \(b\) ), \(\mu\) is a local transition function
(described below), \(k\) is an integer denoting the distinguished
acceptance tape square, and \(A \subseteq \Sigma\) is a set of accepting
symbols. \(M\) is assumed to have a single read-write tape head and a
single two-way infinite tape with tape squares indexed by
\(\mathbb{Z}\).

A configuration of a QTM \(M\) is a triple \((s, h, c)\), where
\(s \in K\) denotes the current state of \(M, h \in \mathbb{Z}\) denotes
the index of the tape square over which the tape head is currently
located, and \(c\) is a map from \(\mathbb{Z}\) to \(\Sigma\) which
describes the contents of the tape. For any configuration of \(M\) it is
assumed that the number of tape squares containing non-blank symbols is
finite. Any configuration of \(M\) in which the tape cell indexed by
\(k\) contains an element of \(A\) is said to be an accepting
configuration and all other configurations are non-accepting
configurations.

The local transition function \(\mu\) is a map

\[\mu: K \times \Sigma \times \Sigma \times K \times\{L, R\} \longrightarrow \mathbb{C}\]

which describes the evolution of \(M\). Suppose that, at some particular
time, the current state of \(M\) is \(s\) and the tape head of \(M\) is
located above a tape square which contains the symbol \(\tau\). Then for
every triple
\(\left(\tau^{\prime}, s^{\prime}, d\right) \in \Sigma \times K \times\{L, R\}, M\)
will write the symbol \(\tau^{\prime}\) in the currently scanned tape
square, change internal state to \(s^{\prime}\) and move the tape head
in direction \(d\) with amplitude
\(\mu\left(s, \tau, \tau^{\prime}, s^{\prime}, d\right)\)

As in the 1d-QCA case, the computation behaves as if all of these
transitions occur simultaneously, so that after some number of steps
\(M\) will have traversed many different computation paths
simultaneously, each with an associated amplitude. The amplitude
associated with each path is defined to be the product of the amplitudes
of the transitions along that path. After \(l\) steps, say, \(M\) will
be in a linear superposition of configurations, and the amplitude
associated with each configuration will be the sum of the amplitudes
along all paths of length \(l\) to that configuration.

If observed while in a superposition of configurations, \(M\) will
randomly choose a single configuration which will be seen; again the
probability that a given configuration will be observed is the absolute
square of the amplitude associated with that configuration. Allowed
local transition functions \(\mu\) must be restricted to those that
guarantee that the sum of these probabilities will be one for any given
superposition. Such a QTM is said to be well-formed. (Write \(M \in\)
QTM whenever \(M\) is a well-formed quantum Turing machine.) We state,
without proof, the following theorem, due to Bernstein and Vazirani,
which allows us to characterize well-formed quantum Turing machines in
terms of their local transition functions:

Theorem 4.1 (Bernstein \& Vazirani) Given any
\(M=(K, \Sigma, \mu, k, A)\) as above, \(M \in Q T M\) if and only if
(i)
\(\forall\left(s_{1}, \tau_{1}\right),\left(s_{2}, \tau_{2}\right) \in K \times \Sigma\)
:

\[\begin{array}{r}
\sum_{\xi, s, d} \mu\left(s_{1}, \tau_{1}, \xi, s, d\right) \overline{\mu\left(s_{2}, \tau_{2}, \xi, s, d\right)} \\
= \begin{cases}1 & \text { if }\left(s_{1}, \tau_{1}\right)=\left(s_{2}, \tau_{2}\right) \\
0 & \text { if }\left(s_{1}, \tau_{1}\right) \neq\left(s_{2}, \tau_{2}\right)\end{cases}
\end{array}\]

and (ii)
\(\forall\left(s_{1}, \tau_{1}, \xi_{1}\right),\left(s_{2}, \tau_{2}, \xi_{2}\right) \in K \times \Sigma \times \Sigma\)
:

\[\sum_{s} \mu\left(s_{1}, \tau_{1}, \xi_{1}, s, L\right) \overline{\mu\left(s_{2}, \tau_{2}, \xi_{2}, s, R\right)}=0 .\]

\hypertarget{equivalence-of-qtm-and-1d-pqca-models}{%
\section{Equivalence of QTM and 1d-PQCA
Models}\label{equivalence-of-qtm-and-1d-pqca-models}}

We now show that for any \(M_{t m} \in \mathrm{QTM}\), there is an
\(M_{c a} \in 1 \mathrm{~d}-\mathrm{PQCA}\) which simulates \(M_{t m}\)
with constant slowdown, and similarly for any
\(M_{c a} \in 1 \mathrm{~d}-\mathrm{PQCA}\) there is an
\(M_{t m} \in \mathrm{QTM}\) which simulates \(M_{c a}\) with linear
slowdown.

\hypertarget{d-pqca-simulation-of-a-qtm}{%
\subsection{1d-PQCA Simulation of a
QTM}\label{d-pqca-simulation-of-a-qtm}}

Let
\(M_{t m}=\left(K, \Sigma, \mu, k_{t m}, A_{t m}\right) \in \mathrm{QTM}\)
and let \(M_{c a}=\left(Q, \Lambda, k_{c a}, A_{c a}\right) \in 1\)
d-PQCA. We will say that \(M_{c a}\) simulates \(M_{t m}\) if and only
if there exists a lineartime computable function
\(T: \mathcal{C}\left(M_{t m}\right) \rightarrow \mathcal{C}\left(M_{c a}\right)\)
and a function
\(f: \mathbb{Z}^{+} \times \mathbb{Z}^{+} \rightarrow \mathbb{Z}^{+}\)such
that for any given configuration
\(a \in \mathcal{C}\left(M_{t m}\right)\), the probability that
\(M_{t m}\) accepts \(a\) after \(t\) steps is equal to the probability
that \(M_{c a}\) accepts \(T(a)\) after \(f(t,|T(a)|)\) steps.

Without loss of generality, we can assume that the QTM which is to be
simulated has the property that any given state is only entered while
the tape head is moving in one direction, due to the following lemma of
Bernstein and Vazirani.

Lemma 5.1 (Bernstein \& Vazirani) Given any \(M \in Q T M\), there
exists an \(M^{\prime} \in Q T M\) which simulates \(M\) with constant
slowdown, and which satisfies the following property: (P1)
\(M^{\prime}\) enters each state while moving in exactly one direction,
i.e. if
\(\mu^{\prime}\left(s_{1}, \tau_{1}, \xi_{1}, s_{1}^{\prime}, d_{1}\right)\)
and
\(\mu^{\prime}\left(s_{2}, \tau_{2}, \xi_{2}, s_{2}^{\prime}, d_{2}\right)\)
are both nonzero, then \(d_{1}=d_{2}\).

Thus, given \(M_{t m}=\left(K, \Sigma, \mu, k_{t m}, A_{t m}\right)\)
satisfying \((P 1), K\) can be partitioned into two sets: \(K_{l}\) and
\(K_{r}\), such that \(M\) enters states in \(K_{l}\) only when moving
left, and enters states in \(K_{r}\) only when moving right.

We now define a
\(1 \mathrm{~d}-\mathrm{PQCA} M_{c a}=\left(Q, \Lambda, k_{c a}, A_{c a}\right)\)
which will simulate \(M_{t m}\). Define
\(Q=Q_{l} \times Q_{m} \times Q_{r}\) with

\[Q_{l}=K_{l} \cup\{\#\}, \quad Q_{m}=\Sigma, \quad Q_{r}=K_{r} \cup\{\#\} .\]

The quiescent element of \(Q\) is defined to be (\#, b, \#), where \(b\)
is the blank symbol of \(M_{t m}\). Define \(\Lambda\) as follows:

(i) for each
\(\left(s_{1}, \tau_{1}\right),\left(s_{2}, \tau_{2}\right) \in K_{l} \times \Sigma\)
let

\[\lambda\left(\left(s_{1}, \tau_{1}, \#\right),\left(s_{2}, \tau_{2}, \#\right)\right)=\mu\left(s_{1}, \tau_{1}, \tau_{2}, s_{2}, L\right),\]

(ii) for each
\(\left(s_{1}, \tau_{1}\right) \in K_{l} \times \Sigma,\left(s_{2}, \tau_{2}\right) \in K_{r} \times \Sigma\)
let

\[\begin{aligned}
& \lambda\left(\left(s_{1}, \tau_{1}, \#\right),\left(\#, \tau_{2}, s_{2}\right)\right)=\mu\left(s_{1}, \tau_{1}, \tau_{2}, s_{2}, R\right), \\
& \lambda\left(\left(\#, \tau_{2}, s_{2}\right),\left(s_{1}, \tau_{1}, \#\right)\right)=\mu\left(s_{2}, \tau_{2}, \tau_{1}, s_{1}, L\right),
\end{aligned}\]

(iii) for each
\(\left(s_{1}, \tau_{1}\right),\left(s_{2}, \tau_{2}\right) \in K_{r} \times \Sigma\)
let

\[\lambda\left(\left(\#, \tau_{1}, s_{1}\right),\left(\#, \tau_{2}, s_{2}\right)\right)=\mu\left(s_{1}, \tau_{1}, \tau_{2}, s_{2}, R\right),\]

(iv) for any \(q_{1}, q_{2} \in Q\) for which
\(\lambda\left(q_{1}, q_{2}\right)\) has not already been defined in (i)
- (iii), let

\[\lambda\left(q_{1}, q_{2}\right)= \begin{cases}1 & \text { if } q_{1}=q_{2} \\ 0 & \text { otherwise }\end{cases}\]

Let \(k_{c a}=k_{t m}\) and define

\[A_{c a}=\left\{\left(q_{l}, q_{m}, q_{r}\right) \in Q \mid q_{m} \in A_{t m}\right\} .\]

A straightforward application of Theorem 3.1 and Theorem 4.1 yields the
following

Lemma 5.2 If \(M_{t m} \in Q T M\) then \(M_{c a} \in 1 d-P Q C A\).

For \((s, h, c) \in \mathcal{C}\left(M_{t m}\right)\), define
\(T((s, h, c)) \in \mathcal{C}\left(M_{c a}\right)\) as follows. For
each \(n \in \mathbb{Z}\) let

\[\begin{aligned}
& T((s, h, c))(n) \\
& = \begin{cases}(\#, c(n), \#) & \text { if } s \in K_{l} \text { and } n \neq h+1 \\
(s, c(n), \#) & \text { if } s \in K_{l} \text { and } n=h+1 \\
(\#, c(n), \#) & \text { if } s \in K_{r} \text { and } n \neq h-1 \\
(\#, c(n), s) & \text { if } s \in K_{r} \text { and } n=h-1 .\end{cases}
\end{aligned}\]

Lemma 5.3 For any \(n \in \mathbb{Z}^{+}\), the probability that
\(M_{t m}\) accepts \((s, h, c)\) after \(n\) steps is equal to the
probability that \(M_{c a}\) accepts \(T((s, h, c))\) after \(n\) steps.

Proof. {[}Sketch{]} For any configuration \(a\) of \(M_{t m}\), the
symbol contained in the tape square indexed by \(n\) is equal to the
contents of the middle subcell of the cell indexed by \(n\) in \(T(a)\).
Recall that the evolution of a 1d-PQCA may be decomposed into two steps:
applying the permutation \(\sigma\) to the current configuration, then
transforming each cell according to \(\lambda\). Now, for each
\(n, \sigma T((s, h, c))(n)\) will have have a non-\# state in its left
or right subcell exactly when \(h=n\), thereby simulating the presence
of the tape head at this location. By inspection of the definition of
\(\lambda\), it is clear that \(M_{c a}\) will evolve in accordance with
\(M_{t m}\).

Together, Lemma 5.1, Lemma 5.2 and Lemma 5.3 give us

Theorem 5.1 Given any \(M_{t m} \in Q T M\) there is an
\(M_{c a} \in 1 d-P Q C A\) which simulates \(M_{t m}\) with constant
slowdown.

\hypertarget{qtm-simulation-of-a-1d-pqca}{%
\subsection{QTM Simulation of a
1d-PQCA}\label{qtm-simulation-of-a-1d-pqca}}

Again, let
\(M_{t m}=\left(K, \Sigma, \mu, k_{t m}, A_{t m}\right) \in \mathrm{QTM}\)
and let
\(M_{c a}=\left(Q, \Lambda, k_{c a}, A_{c a}\right) \in 1 \mathrm{~d}-\mathrm{PQCA}\).
Then \(M_{t m}\) simulates \(M_{c a}\) if and only if there exists a
linear-time computable function
\(T: \mathcal{C}\left(M_{c a}\right) \rightarrow \mathcal{C}\left(M_{t m}\right)\)
and a function
\(f: \mathbb{Z}^{+} \times \mathbb{Z}^{+} \rightarrow \mathbb{Z}^{+}\)such
that for any given configuration
\(a \in \mathcal{C}\left(M_{c a}\right)\), the probability that
\(M_{c a}\) accepts \(a\) after \(t\) steps is equal to the probability
that \(M_{t m}\) accepts \(T(a)\) after \(f(t,|T(a)|)\) steps.

Given
\(M_{c a}=\left(Q, \Lambda, k_{c a}, A_{c a}\right) \in 1 \mathrm{~d}-\mathrm{PQCA}\)
we will now define
\(M_{t m}=\left(K, \Sigma, \mu, k_{t m}, A_{t m}\right) \in \mathrm{QTM}\)
which will simulate \(M_{c a}\).

Each tape square of \(M_{t m}\) will represent the cell of \(M_{c a}\)
with the same index. Define

\[\Sigma=\left(Q_{l} \times Q_{m} \times Q_{r}\right) \cup\left(Q_{l} \times Q_{m} \times Q_{r}\right)^{\star}\]

where \(\left(Q_{l} \times Q_{m} \times Q_{r}\right)^{\star}\) is a copy
of \(\left(Q_{l} \times Q_{m} \times Q_{r}\right)\), so that
\(\left(p_{l}, p_{m}, p_{r}\right)^{\star} \in\left(Q_{l} \times Q_{m} \times Q_{r}\right)^{\star}\)
whenever
\(\left(p_{l}, p_{m}, p_{r}\right) \in\left(Q_{l} \times Q_{m} \times Q_{r}\right)\).
( \(M_{t m}\) will need to simulate only a finite portion of
\(M_{c a}\), since we are only concerned with finite configurations of
\(M_{c a}\), so tape squares containing a symbol in
\(\left(Q_{l} \times Q_{m} \times Q_{r}\right)^{\star}\) will be used to
mark the leftmost and rightmost cells of \(M_{c a}\) being simulated at
that time.) Let \(\epsilon\) be the blank symbol of \(M_{t m}\).

Define

\[K=Q_{l} \times\left\{s_{0}, s_{1}, s_{1}^{\star}, s_{2}, s_{2}^{\star}, s_{3}\right\} \times Q_{r} .\]

Each state of \(M_{t m}\) will consist of one element from \(Q_{l}\) and
one element from \(Q_{r}\) (in order to "move" states from one tape
square to another when performing the permutation \(\sigma\), as
described in section 3) and one element from
\(\left\{s_{0}, s_{1}, s_{1}^{\star}, s_{2}, s_{2}^{\star}, s_{3}\right\}\).

The simulation of each step of \(M_{c a}\) takes place in three stages.
The first stage reversibly performs the permutation \(\sigma\) on the
configuration of \(M_{c a}\) currently represented by the contents of
the tape. In order to do this, let \(\mu\) take the following values:
for each \((l, r) \in\)
\(Q_{l} \times Q_{r},\left(p_{l}, p_{m}, p_{r}\right),\left(q_{l}, q_{m}, q_{r}\right) \in Q_{l} \times Q_{m} \times Q_{r}\),
define

\[\begin{aligned}
& \mu\left(\left(l, s_{0}, r\right),\left(p_{l}, p_{m}, p_{r}\right)^{\star},\left(p_{l}, p_{m}, r\right)^{\star},\left(l, s_{1}, p_{r}\right), R\right)=1 \\
& \mu\left(\left(l, s_{1}, r\right),\left(p_{l}, p_{m}, p_{r}\right),\left(p_{l}, p_{m}, r\right),\left(l, s_{1}, p_{r}\right), R\right)=1 \\
& \mu\left(\left(l, s_{1}, r\right),\left(p_{l}, p_{m}, p_{r}\right)^{\star},\left(p_{l}, p_{m}, r\right),\left(l, s_{1}^{\star}, p_{r}\right), R\right)=1 \\
& \mu\left(\left(l, s_{1}^{\star}, r\right),\left(p_{l}, p_{m}, p_{r}\right)^{\star},\left(p_{l}, p_{m}, r\right)^{\star},\left(l, s_{1}^{\star}, p_{r}\right), R\right)=1 \\
& \mu\left(\left(l, s_{1}^{\star}, r\right),\left(p_{l}, p_{m}, p_{r}\right),\left(l, p_{m}, r\right)^{\star},\left(p_{l}, s_{2}, p_{r}\right), L\right)=1 \\
& \mu\left(\left(l, s_{2}, r\right),\left(p_{l}, p_{m}, p_{r}\right),\left(l, p_{m}, p_{r}\right),\left(p_{l}, s_{2}, r\right), L\right)=1 \\
& \mu\left(\left(l, s_{2}, r\right),\left(p_{l}, p_{m}, p_{r}\right)^{\star},\left(l, p_{m}, p_{r}\right),\left(p_{l}, s_{2}^{\star}, r\right), L\right)=1 \\
& \mu\left(\left(l, s_{2}^{\star}, r\right),\left(p_{l}, p_{m}, p_{r}\right)^{\star},\left(l, p_{m}, p_{r}\right)^{\star},\left(p_{l}, s_{2}^{\star}, r\right), L\right)=1
\end{aligned}\]

(Note that not all of these transitions will be used given proper input,
but are added to guarantee wellformedness.) In the second stage of the
simulation, the state contained in each tape square of \(M_{t m}\) is
transformed in accordance with \(\lambda\), so for each \((l, r) \in\)
\(Q_{l} \times Q_{r},\left(p_{l}, p_{m}, p_{r}\right),\left(q_{l}, q_{m}, q_{r}\right) \in Q_{l} \times Q_{m} \times Q_{r}\),
define

\[\begin{aligned}
& \mu\left(\left(l, s_{2}^{\star}, r\right),\left(p_{l}, p_{m}, p_{r}\right),\right.\left.\left(q_{l}, q_{m}, q_{r}\right)^{\star},\left(p_{l}, s_{3}, r\right), R\right) \\
&=\lambda\left(\left(l, p_{m}, p_{r}\right),\left(q_{l}, q_{m}, q_{r}\right)\right) \\
& \mu\left(\left(l, s_{3}, r\right),\left(p_{l}, p_{m}, p_{r}\right),\right.\left.\left(q_{l}, q_{m}, q_{r}\right),\left(l, s_{3}, r\right), R\right) \\
&\left.=\lambda\left(p_{l}, p_{m}, p_{r}\right),\left(q_{l}, q_{m}, q_{r}\right)\right) \\
& \mu\left(\left(l, s_{3}, r\right),\left(p_{l}, p_{m}, p_{r}\right)^{\star},\right.\left.\left(q_{l}, q_{m}, q_{r}\right)^{\star},\left(l, s_{0}, r\right), L\right) \\
&=\lambda\left(\left(p_{l}, p_{m}, p_{r}\right),\left(q_{l}, q_{m}, q_{r}\right)\right)
\end{aligned}\]

The third stage of the simulation simply returns the tape head to the
tape square representing the (new) leftmost cell of \(M_{c a}\) being
simulated. For each \((l, r) \in\)
\(Q_{l} \times Q_{r},\left(p_{l}, p_{m}, p_{r}\right),\left(q_{l}, q_{m}, q_{r}\right) \in Q_{l} \times Q_{m} \times Q_{r}\),
define

\[\mu\left(\left(l, s_{0}, r\right),\left(p_{l}, p_{m}, p_{r}\right),\left(p_{l}, p_{m}, p_{r}\right),\left(l, s_{0}, r\right), L\right)=1\]

Finally, let \(\mu\) take the value 0 everywhere not defined above.

Let \(k_{t m}=k_{c a}\) and define

\[A_{t m}=\left\{q \mid q \in A_{c a}\right\} \cup\left\{q^{\star} \mid q \in A_{c a}\right\} .\]

The following lemma follows from Theorem 3.1 and Theorem 4.1.

Lemma 5.4 If \(M_{c a} \in 1 d-P Q C A\) then \(M_{t m} \in Q T M\).

Given \(a \in \mathcal{C}\left(M_{c a}\right)\) let \(n_{l}\) and
\(n_{r}\) denote the indices of the leftmost and rightmost non-quiescent
cells of \(a\) respectively. (If \(n_{l}=n_{r}\) then set
\(n_{r}=n_{l}+1\). If all cells of \(a\) are quiescent, then choose
\(n_{l}\) arbitrarily and set \(n_{r}=n_{l}+1\).) Define
\((s, h, c)=T(a)\) as follows: let
\(s=\left(\epsilon, s_{0}, \epsilon\right)\), let \(h=n_{l}\), and let
\(c: \mathbb{Z} \rightarrow \Sigma\) be defined as

\[c(n)= \begin{cases}a(n) & \text { if } n \neq n_{l} \text { and } n \neq n_{r} \\ a(n)^{\star} & \text { if } n=n_{l} \text { or } n=n_{r}\end{cases}\]

Define

\[f(t,|c|)=4 t^{2}+4|c| t-t\]

This is the number of steps required for \(M_{t m}\) to simulate \(t\)
steps of \(M_{c a}\) on input \(c\)..

Lemma 5.5 For any \(t \in \mathbb{Z}^{+}\)and
\(c \in \mathcal{C}\left(M_{c a}\right)\), the probability that
\(M_{c a}\) accepts \(c\) after \(t\) steps is equal to the probability
that \(M_{t m}\) accepts \(T(c)\) after \(f(t,|T(c)|)\) steps.

Proof. {[}Sketch{]} Assume that we begin the simulation of each time
step of \(M_{c a}\) with the tape head located over the leftmost cell of
the finite section of \(M_{c a}\) being simulated, and with the current
internal state of \(M_{t m}\) equal to
\(\left(\epsilon, s_{0}, \epsilon\right)\).

First, the permutation \(\sigma\) is performed reversibly in two passes
of the tape contents, first moving to the far right then returning to
the left. The finite section of \(M_{c a}\) being simulated will grow by
one cell on each end during this stage, since the right and left starred
symbols are shifted one cell to the right and left respectively. Next,
in a single pass to the right, the state of each tape square is
"quantumly" transformed in accordance with \(\lambda\). When this has
been completed, the tape head is returned to the tape square
representing the (new) leftmost cell of \(M_{c a}\) being simulated.
Since the tape squares outside of the region between the starred symbols
are assumed to contain blanks (representing quiescent cells), the
current internal state of \(M_{t m}\) will again be
\(\left(\epsilon, s_{0}, \epsilon\right)\).

Thus, after four passes of the current tape contents, \(M_{t m}\) will
be in a linear superposition of configurations which corresponds to the
superposition of \(M_{c a}\) after one time step, i.e. each
configuration of \(M_{t m}\) will have the same amplitude as the
configuration in the superposition of \(M_{c a}\) which it represents.
The long-term evolution of \(M_{t m}\) will thus behave exactly as that
of \(M_{c a}\).

By Lemma 5.4 and Lemma 5.5 we have

Theorem 5.2 Given any \(M_{c a} \in 1 d-P Q C A\), there exists
\(M_{t m} \in Q T M\) which simulates \(M_{c a}\) with linear slowdown.

\hypertarget{some-resulting-facts}{%
\section{Some Resulting Facts}\label{some-resulting-facts}}

From the preceeding discussion, we are immediately led to two
straightforward results.

The first result is an observation concerning the construction in
section 5.2. The QTM which results from this construction has the
interesting property that the motion of its tape head is deterministic,
formalized as follows. Let \(M \in \mathrm{QTM}\) and let \(a\) be an
arbitrary configuration of \(M\). If \(M\) is assumed to be in
configuration \(a\) and is run (unobserved) for \(l\) steps and then
observed, then for any integer \(h\) there is some probability that the
tape head of \(M\) will be located over the tape square indexed by
\(h\). We say that \(M\) has deterministic head position if, for all
\(a \in \mathcal{C}(M)\), \(l \in \mathbb{Z}^{+}\)and
\(h \in \mathbb{Z}\), we have that this probability is either zero or
one.

By Theorem 5.1 and Theorem 5.2 we have

Corollary 6.1 For any QTM M, there exists a QTM \(M^{\prime}\) such that
\(M^{\prime}\) simulates \(M\) with linear slowdown, and such that
\(M^{\prime}\) has deterministic head position.

In addition to being deterministic, the head position of the QTM
resulting from the previously mentioned construction is also oblivious;
for any input of a given length, the position of the tape head at a
given time will depend only upon this length.

The second result regards the notion of acceptance and rejection by a
1d-PQCA. Consider

\[M=(Q, \Lambda, k, A) \in 1 \mathrm{~d}-\mathrm{PQCA} .\]

Recall that for any \(a \in \mathcal{C}\) and
\(l \in \mathbb{Z}^{+}, M\) accepts \(a\) after \(l\) steps with
probability equal to that of observing an element of \(A\) in the cell
indexed by \(k\) after \(l\) steps. Of course, this probability may
differ greatly with any change in \(l\), so that if we are interested in
whether or not \(M\) recognizes some language
\(\mathcal{L} \subseteq \mathcal{C}(M)\), we must pair some
\(l \in \mathbb{Z}^{+}\)with each input \(a \in \mathcal{C}\) and
observe \(M\) after exactly \(l\) steps have passed. This may seem
unsatisfactory since it implies that the observer must carefully clock
the machine and observe it at precisely the right moment, or else the
computation may be invalid. However, we see that this is a reasonable
definition of acceptance/rejection, due to the following

Theorem 6.1 For any \(M \in 1 d-P Q C A\) there exists an
\(M^{\prime} \in 1 d-P Q C A\) so that for any \(a \in \mathcal{C}(M)\)
and \(l \in \mathbb{Z}^{+}\) there exists an
\(a^{\prime} \in \mathcal{C}\left(M^{\prime}\right)\) (computable in
\(\mathcal{O}(l+|a|)\) steps) such that the probability that \(M\)
accepts a after \(l\) steps is equal to the probability that
\(M^{\prime}\) accepts a' after \(l^{\prime}\) steps for any
\(l^{\prime} \geq l\). Proof. Suppose that a 1d-PQCA
\(M=(Q, \Lambda, k, A)\), with quiescent state \(\epsilon\), is given.

First, define \(M_{t}=\left(Q_{t}, \Lambda_{t}, k, A_{t}\right)\) as
follows. Let

\[Q_{t}=\{0,1\} \times\{0,1\} \times\{0\}\]

(with \(\epsilon_{t}=(0,0,0)\) as the quiescent state), let
\(\Lambda_{t}\) be the identity matrix and let \(A_{t}=\{(1,1,0)\}\).
The action of \(M_{t}\) on any configuration is identical to the
permutation \(\sigma\).

For any given \(l \in \mathbb{Z}^{+}\), if we define
\(a_{t} \in \mathcal{C}\left(M_{t}\right)\) as

\[a_{t}(n)= \begin{cases}(0,1,0) & \text { if } n=k \\ (1,0,0) & \text { if } n=k+l \\ (0,0,0) & \text { otherwise, }\end{cases}\]

then on input \(a_{t}\), the cell of \(M_{t}\) indexed by \(k\) will be
in the state \((1,1,0)\) after exactly \(l\) steps have passed, and at
no other time will any cell contain this state.

Now define \(M_{1}=\left(Q_{1}, \Lambda_{1}, k, A_{1}\right)\) as
follows. Let \(Q_{1}=Q \times Q_{t}\) (with quiescent state
\(\epsilon_{1}=\left(\epsilon, \epsilon_{t}\right)\), let
\(A_{1}=A \times A_{t}\) and let \(\Lambda_{1}\) be as defined by

\[\lambda_{1}\left(\left(p, p_{t}\right),\left(q, q_{t}\right)\right)=\lambda(p, q) \lambda_{t}\left(p_{t}, q_{t}\right)\]

for \(\left(p, p_{t}\right),\left(q, q_{t}\right) \in Q \times Q_{t}\).
Clearly \(M_{1} \in 1 \mathrm{~d}-\mathrm{PQCA}\). \(M_{1}\) can be
viewed as \(M\) and \(M_{t}\) "stacked" on top of one another, evolving
independently.

For any given \(a \in \mathcal{C}(M)\) and \(l \in \mathbb{Z}^{+}\), if
we define \(a_{1} \in \mathcal{C}\left(M_{1}\right)\) as

\[a_{1}(n)=\left(a(n), a_{t}(n)\right),\]

where \(a_{t}\) is as defined above, then the probability that \(M_{1}\)
accepts \(a_{1}\) after \(l\) steps is the same as the probability that
\(M\) accepts \(a\) after \(l\) steps. Furthermore, the cell of
\(M_{1}\) indexed by \(k\) will not contain an accepting state at any
other time, and at no time will any other cell contain an accepting
state.

Finally, define \(Q_{a}=\{0\} \times\{0,1\} \times\{0\}\), and define
\(M^{\prime}=\left(Q^{\prime}, \Lambda^{\prime}, k, A^{\prime}\right)\)
as follows. Let \(Q^{\prime}=Q_{1} \times Q_{a}\), let
\(A^{\prime}=Q_{1} \times\{(0,1,0)\}\) and define \(\Lambda^{\prime}\)
as

\[\begin{aligned}
& \lambda^{\prime}\left(\left(p_{1}, p_{a}\right),\left(q_{1}, q_{a}\right)\right) \\
& \quad= \begin{cases}\lambda_{1}\left(p_{1}, q_{1}\right) & \text { if } q_{1} \notin A_{1} \text { and } p_{a}=q_{a} \\
0 & \text { or } q_{1} \in A_{1} \text { and } p_{a} \neq q_{a}\end{cases}
\end{aligned}\]

for every
\(\left(p_{1}, p_{a}\right),\left(q_{1}, q_{a}\right) \in Q_{1} \times Q_{a}\).
Since \(\Lambda_{1}\) is unitary, we must have that \(\Lambda^{\prime}\)
is unitary, so that \(M^{\prime} \in 1\) d-PQCA (with quiescent state
\(\left(\epsilon_{1},(0,0,0)\right)\).)

Now, for given \(a \in \mathcal{C}(M)\) and \(l \in \mathbb{Z}^{+}\),
define \(a^{\prime} \in \mathcal{C}\left(M^{\prime}\right)\) as

\[a^{\prime}(n)=\left(a_{1}(n),(0,0,0)\right),\]

where \(a_{1}\) is as defined above. Consider the evolution of
\(M^{\prime}\) on input \(a^{\prime}\). By the definition of
\(\Lambda^{\prime}\), we see that any state
\(\left(p_{1}, p_{a}\right)\) will be transformed (with nonzero
amplitude) into state \(\left(q_{1}, q_{a}\right)\) with
\(q_{a} \neq p_{a}\) (toggling acceptance/rejection) exactly when
\(q_{1} \in A_{1}\). But if any cell of \(M^{\prime}\) ever contains a
state in \(A_{1} \times Q_{a}\), then we know that this must be the cell
indexed by \(k\) and exactly \(l\) steps must have passed (by the
property of \(M_{1}\) on input \(a_{1}\) mentioned above.) Thus, such a
transformation will occur at most one time on any given computation
path, and only on those paths which correspond to paths of \(M_{1}\)
which accept \(a_{1}\) after \(l\) steps.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

Many thanks to Eric Bach for his valuable insights and advice, and for
introducing me to quantum computing. Thanks also to Anne Condon and
Walter Ludwig for their helpful comments and suggestions, and to
Christoph Drr for pointing out an error in an earlier version of this
paper.

\hypertarget{references}{%
\section{References}\label{references}}

{[}1{]} R. Feynman, Simulating Physics with Computers, International
Journal of Theoretical Physics, Vol. 21, nos. 6/7 (1982) 467-488.

{[}2{]} D. Deutsch, Quantum Theory, the Church-Turing Principle and the
Universal Quantum Computer, Proc. R. Soc. Lond., Vol. A400 (1985)
97-117.

{[}3{]} E. Bernstein and U. Vazirani, Quantum Complexity Theory, Proc.
25th Ann. ACM Symp. on Theory of Computing (1993) 11-20.

{[}4{]} A. Yao, Quantum Circuit Complexity, Proc. 34th Ann. Symp.
Foundations of Computer Science (1993) 352-361.

{[}5{]} P. Shor, Algorithms for Quantum Computation: Discrete Logarithms
and Factoring, Proc. 35th Ann. Symp. Foundations of Computer Science
(1994) 124-134.

{[}6{]} K. Morita and M. Harao, Computation Universality of
One-Dimensional Reversible (Injective) Cellular Automata, Transactions
of the IEICE, Vol. E 72 (1989) 758-762.

{[}7{]} R. Cooke, Infinite Matrices and Sequence Spaces, Macmillan and
Co., London, 1950.

{[}8{]} N. Young, An Introduction to Hilbert Space, Cambridge University
Press, Cambridge, 1988.

\end{document}
